---
title: "DE analysis on A. thaliana BIN2/RAPTOR1B mutant plants using PoissonSeq"
author: "Christian Montes"
date: "May 12, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We first load the required libraries
```{r message=FALSE}
shhh <- suppressWarnings # It's a library, so shhh!
shhh(library(PoissonSeq))
shhh(library(ggbiplot))
shhh(library(edgeR))
shhh(library(openxlsx))
shhh(library(EnhancedVolcano))
shhh(library(tidyverse))
shhh(library(knitr))
```

## Here is where you provide basic information for the analysis: counts file, number of replicates, and set qvalue/FC cutoff
```{r, message=FALSE}
metadata <- read_csv("metadata.csv")
read_csv("read_counts_BR.csv") %>%
#  select(gname, 1:17) %>%
  `colnames<-`(c("gname", paste(metadata$name,metadata$`Bio Rep`, sep = ""))) %>%
  column_to_rownames(var = "gname") -> counts
qval <- 0.05
FC <- 2
DE_list <- "DE_pairs.csv" #leave blank if you don't have a list of specific pairwise comparisons to test, otherwise provide filename
nreps <- 4
multicomp <- "no"

```

# Let's see how samples cluster (in a PCA plot)

```{r, echo=FALSE}
#perform PCA
pcaresults = prcomp(t(floor(log2(counts+1))))

#plot PCA
g <- ggbiplot(pcaresults, groups = factor(unlist(metadata$name)),
              labels=colnames(counts),
              var.axes=FALSE,labels.size=3,ellipse=TRUE) +
  theme(text = element_text(size=14), aspect.ratio = 1) +
  ggtitle(label = "All samples included") +
  xlim(-2.5, 2.5)
print(g)
```

Let's see how clustering looks without sample 'bin2T_M2'

```{r, echo=FALSE}
#get rid of bad rep
pcaresultsnoBadRep = prcomp(t(floor(log2(counts+1)[,c(1,3:32)])))
#plot PCA
g <- ggbiplot(pcaresultsnoBadRep, groups = factor(unlist(metadata[c(1,3:32),2])),
              labels=colnames(counts[c(1,3:32)]),
              var.axes=F,labels.size=3,ellipse=TRUE) +
    theme(text = element_text(size=14), aspect.ratio=1) + 
    ggtitle(label = "Sample bin2T_M2 removed")
print(g)
```

Clustering improves when we remove sample bin2T_M2.

#Performing PoissonSeq DE analysis

## If we need, We can first do multiclass comparisons to get the genes DE in at least one comparison. If now, we skip this step.
```{r}
counts <- counts[,c(1,3:32)]
if (multicomp == "yes") {
  yall=c(rep(1, times = 3),c(rep(2:8, each = 4)))
  pseq<- PS.Main(dat=list(n=counts,y=yall,type="multiclass",pair=FALSE,gname=row.names(counts)))
  write.csv(pseq,'PSeq_first_multiclass_comp.csv')
  mygenes <- pseq[pseq$fdr<0.05,]
  write.csv(mygenes,'PSeq_first_multiclass_comp_q0.05.csv')
} else { }

```

## We now perform pairwise comparisons using only the significant genes from previous multiclass comparison (This helps increase the power).

We want to get the TMM-normalized count values to use as fold change value later
```{r}
counts = counts[order(row.names(counts)),]
norm.factors = calcNormFactors(counts,method="TMM")
tmm = equalizeLibSizes(counts,norm.factors=norm.factors)$pseudo.counts
write.csv(tmm,"TMM_all.csv")
```

We extract only the genes significant from our multiclass comparison
```{r}
if (multicomp == "yes") {
  mygenes = mygenes[order(mygenes$gname),]
  mycounts = counts[row.names(counts)%in%mygenes$gname,]
} else {
  mycounts = counts
}
```

And, either make a list of all pairwise comparisons, or provide a file with specific comparison to test
```{r, message=FALSE}
if (DE_list == "") {
  mysamples = unique(metadata$name)
  mysamples = mysamples[!grepl("Ref",mysamples,ignore.case=TRUE)]
  comps = rep("",choose(length(mysamples),2))
  currentrow = 1
  for (i in 1:(length(mysamples)-1)){
    for (j in (i+1):length(mysamples)){
      comps[currentrow] = paste(mysamples[i],"_vs_",mysamples[j],sep="")
      currentrow = currentrow+1
    }
  }
} else {
  comps <- read_csv(DE_list, skip = 1)
  comps$pairs <- paste(comps$control, "_vs_", comps$treatment, sep = "")
  comps <- comps$pairs
}
```

## Now we can perform PoissonSeq analysis
```{r}
pseqdata = mycounts
newwb <- createWorkbook()
newwb2 <- createWorkbook()
for (i in 1:length(comps)){
  #get the intensities for this comparison
  sepcomps = strsplit(comps[i],"_vs_")
  intensities1 = pseqdata[,grepl(sepcomps[[1]][1],colnames(pseqdata))]
  intensities2 = pseqdata[,grepl(sepcomps[[1]][2],colnames(pseqdata))]

  #make indicator variable y
  y= c(rep(1,dim(intensities1)[2]),rep(2,dim(intensities2)[2]))
  #perform PSeq
  pdata = data.frame(intensities1,intensities2)
  pseq<- PS.Main(dat=list(n=pdata,y=y,type="twoclass",pair=FALSE,gname=row.names(pdata)))
  #get the actual fc from TMM-normalized counts
  TMMint1 <- tmm[,grepl(sepcomps[[1]][1],colnames(tmm))]
  TMMint2 <- tmm[,grepl(sepcomps[[1]][2],colnames(tmm))]

  TMMint1 = TMMint1[order(row.names(TMMint1)),]
  TMMint2 = TMMint2[order(row.names(TMMint2)),]
  myFC = data.frame(rowMeans(TMMint2[,])/rowMeans(TMMint1[,]),row.names=row.names(TMMint1))
  myFC <- rownames_to_column(myFC, var = "gname")
  pseq %<>% inner_join(myFC, by="gname") %>%
    mutate(log2FC = log2(rowMeans.TMMint2......rowMeans.TMMint1.....)) %>%
    arrange(gname)
  
  myresults = data.frame(pseq[,c(1:5,8)],TMMint1[row.names(TMMint1)%in%pseq$gname,],
                         TMMint2[row.names(TMMint2)%in%pseq$gname,])
  #save
  addWorksheet(wb = newwb2, sheetName = comps[i], gridLines = TRUE)
  writeDataTable(wb=newwb2, sheet=comps[i],x=myresults,tableStyle="none",
                 rowNames=FALSE,withFilter=FALSE,
                 bandedRows=FALSE,bandedCols=FALSE)
  
  #make volcano plot
  v <- EnhancedVolcano(pseq,rownames(pseq),'log2FC','fdr',
                       pointSize=1,labSize=0,FCcutoff=1,pCutoff=qval,
                       title=paste(comps[i],"(",sum(pseq$fdr<qval),")",sep=""),
                       xlim = c(-3,3), ylim = c(0,5), col=c('grey30','grey60','royalblue','red2'),
                       legendLabels=c('NS','FC>2, q>0.05 (NS)','FC<2, q<0.05 (S)','FC>2, q<0.05 (S)'),
                       legendLabSize=10, ylab = bquote(~-Log[10]~italic(q)))
  
  assign(paste("volc",sepcomps[[1]][2],"vs", sepcomps[[1]][1], sep = "_"), v)
  
  
  #make qvalue histogram
  h <- ggplot(pseq, aes(pval)) + 
    geom_histogram(bins = 200, fill = "white", color = "black") + 
    geom_hline(yintercept = mean(hist(pseq$pval, breaks = 100, plot = FALSE)$counts[26:100])) +
    ggtitle(paste(sepcomps[[1]][2],"vs", sepcomps[[1]][1],"p-value distribution", sep = " "))
  assign(paste("pval",sepcomps[[1]][2],"vs", sepcomps[[1]][1], sep = "_"), h)
 
  #get differentially expressed genes and save
  mypros = subset(myresults, myresults$fdr<qval)
  mypros <- subset(mypros, log2FC >= log2(FC) | log2FC <= -log2(FC))
  #save
  addWorksheet(wb = newwb, sheetName = comps[i], gridLines = TRUE)
  writeDataTable(wb=newwb, sheet=comps[i],x=mypros,tableStyle="none",
                 rowNames=FALSE,withFilter=FALSE,
                 bandedRows=FALSE,bandedCols=FALSE)
  if (i == 1) {
    data.frame("pair","Sig", "up", "down") %>%
    `colnames<-`(c("pair","Sig", "up", "down")) %>%
    mutate(pair = comps[i], Sig = nrow(mypros)-1, up = sum(mypros$log2FC>0),
           down = sum(mypros$log2FC<0)) -> de_data
  } else {
    data.frame("pair","Sig", "up", "down") %>%
    `colnames<-`(c("pair","Sig", "up", "down")) %>%
    mutate(pair = comps[i], Sig = nrow(mypros)-1, up = sum(mypros$log2FC>0),
           down = sum(mypros$log2FC<0)) %>%
      bind_rows(de_data) -> de_data
  }
  
}
colnames(de_data) <- c("DE pair","Sig. DE genes (0.05 FDR, FC2)", "Upreg", "Downreg")
plotnames <- ls(knitr::knit_global(),pattern = "^pval.*$")
volcnames <- ls(knitr::knit_global(),pattern = "^volc.*$")
```

## PoissonSeq DE analysis summary table

```{r, echo = FALSE, results = 'asis'}
kable(de_data)
```

## Visualizing the results

What does the test p-value distribution look like?
```{r echo=FALSE, message=FALSE, warning=FALSE}
mget(plotnames)
```

Let's do some Volcano plots!

```{r, echo=FALSE, message=FALSE}
# make a volcano plot
mget(volcnames)
```

# Finally, we export the results into an Excel file
```{r}
saveWorkbook(newwb, paste("Pseq_all_comps_q",qval,"_FC", FC,".xlsx",sep=""),overwrite=TRUE)
saveWorkbook(newwb2, "Pseq_all_comps.xlsx",overwrite=TRUE)
message("Finished!")
```
